Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-04-29 text/paper2/ms.tex

\begin{abstract}
% Context
The dark-matter halo of the Milky Way is expected to be triaxial on large scales
  and filled with substructure.
It is hoped that streams or shells of stars produced by tidal disruption
  of smaller galaxies or globular clusters will provide very precise measures of 
  the gravitational potential, 
  to test these predictions.
% Aims
We develop a statistically justified method for inferring the gravitational potential
  using precise measurements of stars in tidal streams
  based on the idea that these stars
  were close in phase space at some point in the past.
Our method is flexible enough to adapt to any form for the Galactic potential.
In particular, it works in phase-space rather than action-space and hence relies neither on
our ability to derive actions nor on the 
assumption of integrability of the potential;
  we do not know that any good approximation to the large-scale 
  Milky Way potential is integrable.
% Methods: 
Our model is probabilistic, with a likelihood function and priors on the parameters.
  These parameters include the present-day phase-space position for each star
  and the progenitor, and thus the method can properly account for both finite observational uncertainties 
  and missing data dimensions.
We test our method on synthetic data sets generated from an N-body simulation
of satellite disruption in a multi-component Milky Way including a triaxial dark matter halo. 
The data sets were generated with
  observational uncertainties to mimic current and near-future surveys of various stars.
% Results:
We find that even with a small number (8) of well-measured stream stars,
  we can infer properties of the triaxial potential ???? with accuracies of order ??? percent ??? and
  missing phase-space coordinates of the stars ??? with accuracies of order ????.
The method generalizes to encompass mixtures of independent debris structures,
  or situations in which star membership in structures is unknown.
% [Get more quantitative here?]
% Conclusions: 
\end{abstract}

@@

% TODO: audit all uses of the word 'model'

@@

Motivated by the strengths of previous work, we identify a minimum set of issues that any approximate stream model or potential recovery method should address (see Section~\ref{sec:discussion} for a more detailed discussion of these points in the context of this work):
\begin{enumerate}
	\item \textbf{Observational uncertainties:} The known debris structures are $\gtrsim$10 kpc from the Sun, where distance and proper motion measurement errors are significant. Thus it is critical for any method that uses tidal debris to incorporate observational uncertainties and missing dimensions in a consistent and justified way. 
	\item \textbf{Form of the potential:} There is large uncertainty in the radial profile, shape, orientation, and graininess of the outer halo and the constancy of these parameters over distance. Properly describing the potential may require non-parametric techniques or complex analytic functions so the method should not rely on the existence of or ability to compute simple conserved orbital properties, such as actions.
	\item \textbf{Multiple debris structures:} Near-future photometric surveys such as \gaia\, and the \project{LSST} will likely discover many new streams and kinematic associations of stars. Potential recovery methods should be able to simultaneously use multiple streams and incorporate other dynamical constraints.
	\item \textbf{Comparing models to data:} Matching generated streams to observed stellar densities is difficult; models that rely on this must account for observational biases, internal properties of the progenitor, and background halo stars. [TODO: are we matching densities?]
	\item \textbf{Computational expense:} Full N-body simulations are expensive to run; incorporating an N-body simulation into a likelihood function evaluation and then performing a parameter search would be computationally intensive and, presently, intractable. [TODO: KVJ?] 
\end{enumerate}

@@

In all simulations, the host potential is taken to be a three-component sum of a Miyamoto-Nagai disk \citep{miyamoto75}, Hernquist spheroid , and a triaxial, logarithmic halo \citep[e.g.,][]{law10}:
\begin{align}
	&\Phi_{\rm disk}(R,z) = -\frac{GM_{\rm disk}}{\sqrt{R^2 + (a + \sqrt{z^2 + b^2})^2}}\\
	&\Phi_{\rm spher}(r) = -\frac{GM_{\rm spher}}{r + c}\\
	&\Phi_{\rm halo}(x,y,z) = \vhalo^2 \ln(C_1 x^2 + C_2 y^2 + C_3 xy + (z/q_z)^2 + r_h^2)
\end{align}
where $C_1$, $C_2$, and $C_3$ are combinations of the $x$ and $y$ axis
ratios ($q_1$, $q_2$) and orientation of the halo with respect to the
baryonic disk ($\phi$):
\begin{align}
  C_1 &= \frac{\cos^2\phi}{q_1^2} + \frac{\sin^2\phi}{q_2^2}\\
  C_2 &= \frac{\sin^2\phi}{q_1^2} + \frac{\cos^2\phi}{q_2^2}\\
  C_3 &= 2\sin\phi\cos\phi \left(q_1^{-2} - q_2^{-2}\right).
\end{align}
The total potential then is just
\begin{equation}
	\Phi_{\rm tot} = \Phi_{\rm disk} + \Phi_{\rm spher} + \Phi_{\rm halo}\label{eq:lm10}.
\end{equation}
Figure~\ref{fig:potential} shows equipotential slices in the Galactic X-Z plane (Y=0) and Y-Z plane (X=0) for a few choices of  $q_1$,  $q_z$, and $\phi$ while holding all other parameters fixed, as described in the figure caption. The potential parameters used for the simulations are shown in the ``Truth'' column of Table~\ref{tbl:params}.

@@

In a triaxial potential, the orbital plane of the satellite is not fixed but we still expect there to be \emph{effective} Lagrange points for a spherical satellite system along the line of centers connecting the origin of the parent potential to the satellite --- that is, we expect the stars to be stripped at some characteristic radius near the tidal radius with some dispersion about this point. We proceed by defining a coordinate system relative to the position and velocity of the progenitor, ($\bs{r}_p$,$\bs{v}_p$), rotated into the instantaneous orbital plane defined by $\bs{L} = \bs{r}_p \times \bs{v}_p$. The (time-dependent) basis vectors are given by
\begin{align}
	\hat{\bs{x}}_1 &= \frac{\bs{r}_p}{\|\bs{r}_p\|}\label{eq:x1}\\
	\hat{\bs{x}}_2 &= \frac{\bs{L} \times \hat{\bs{x}}_1}{\|\bs{L} \times \hat{\bs{x}}_1\|} = \hat{\bs{x}}_3 \times \hat{\bs{x}}_1\\
	\hat{\bs{x}}_3 &= \hat{\bs{L}} = \frac{\bs{L}}{\|\bs{L}\|} = \frac{\bs{r}_p \times \bs{v}_p}{\|\bs{r}_p \times \bs{v}_p\|}\label{eq:x3}.
\end{align}
Figure~\ref{fig:lpts_r} shows sections of particle orbits (for particles that are stripped) from the N-body simulations described above projected into this coordinate system. The positions are normalized by the instantaneous tidal radius (Eq.~\ref{eq:tidalradius}) and velocities are normalized by the instantaneous velocity scale (Eq.~\ref{eq:velscale}). Sections of the orbits symmetric around their unbinding times, $\tub$, are shown for each of the four progenitor masses. In these coordinates, the classical Lagrange points would be located at $x_1\approx\pm\rtide,x_2=0,x_3=0$ (illustrated by the intersection of the black unit circles and horizontal lines in Figure~\ref{fig:lpts_r}). Though not centered on the point-mass Lagrange points, the location of and dispersion about the effective Lagrange points in these scaled coordinates remain remarkably consistent across the range of progenitor masses explored. Figure~\ref{fig:lpts_v} shows the velocity orbits of each star also projected into these coordinates and normalized. The dispersion in velocity is well-normalized by the velocity scale of Eq.~\ref{eq:velscale}, though it is clear that the velocity dispersion along $\hat{x}_1$, the radial vector to the satellite position, is larger than that in other dimensions, possibly because of tidal shocking.

@@

@@

To model the true, 6D, present-day position of the star, $\W$, we first transform to cartesian, Galactocentric coordinates where ($\bs{r}_0,\bs{v}_0$) and ($\bs{r}_{\sat,0},\bs{v}_{\sat,0}$) are the position and velocity of the star and progenitor today. The star is taken as having been sampled at $t=\tub$ (the unbinding time) from an isotropic Gaussian centered on one of the effective Lagrange points in position, and an isotropic Gaussian centered on the origin in velocity. The present-day phase-space position is the result of integrating this sample forward from $\tub$ in the parent potential, $\Phi$, whose form is parametrized by the vector $\bs{\Theta}_\Phi$. We assume that once the star becomes ``unbound'' from the satellite the potential of the satellite can be ignored,\footnote{This assumption breaks down for sufficiently large-mass progenitors, somewhere between $\sim10^8-10^9~\msun$.} and thus treat the star as a test particle. The satellite mass enters this prescription through the tidal radius; a changing satellite mass will skew the positions of the effective Lagrange points and velocity scale. We allow for mass-loss by incorporating the initial satellite mass, $m_0$, and a constant mass-loss term, $\dot{m}$ into the model: $m(t) = m_0 - \dot{m}t$. Finally, we add two additional parameters: a constant, global factor, $\Loffset$, that scales the position of the effective Lagrange points relative to the classical tidal radius, and a binary parameter $\tailbit$, that is either $-1$ or $+1$ depending on whether the star is in the leading or trailing tail. To compress notation, we pack all progenitor parameters ($\W_\sat, m_0, \dot{m}, \Loffset$) into the vector $\bs{\Theta}_\sat$, and all star parameters ($\W, \tailbit, \tub$) into the vector $\bs{\Theta}$. The likelihood is then:
\begin{align}
	p(\bs{\Theta} \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi) &= p(\bs{r}_0,\bs{v}_0, \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi)\,\left\vert\J\right\vert_{t=\tub}\label{eq:starlike}\\
	p(\bs{r}_0,\bs{v}_0, \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi) &= \mathcal{N}(\bs{r} \given \bs{r}_\sat - \Loffset\,\tailbit\,\rtide\,\hat{x}_1, \rtide^2)\,\mathcal{N}(\bs{v} \given \bs{v}_\sat, \sigma^2_v)_{t=\tub}
\end{align}
where $\rtide$ is given by Eq.~\ref{eq:tidalradius}, and $\sigma_v$ by Eq.~\ref{eq:velscale}. $\left\vert\J\right\vert = \left\vert\frac{\partial(x,y,z,v_x,v_y,v_z)}{\partial(l,b,d,\mu_l,\mu_b,v_r)}\right\vert$ is the absolute value of the determinant of the Jacobian that defines the transformation from heliocentric, spherical to Galactocentric, cartesian coordinates. All quantities in this expression are evaluated at $t=\tub$. For a single star, the evaluation of the likelihood works as follows: given present-day coordinates for the star and progenitor, potential parameter values ($\bs{\Theta}_\Phi$), and nuisance parameter values ($\tub,\Loffset,\tailbit$),
\begin{enumerate}
	\item transform star and progenitor positions in heliocentric coordinates to Galactocentric, cartesian coordinates;
	\item integrate star and progenitor orbits backwards as test particles in the potential given by $\bs{\Theta}_\Phi$ for the total interaction time, $t_{\rm int}\approx6$~Gyr;
	\item at the timestep closest to $\tub$, transform the particle position to the relative, normalized coordinates defined in Section~\ref{sec:debris};
	\item compute the likelihood given by Equation~\ref{eq:starlike}.
\end{enumerate}
Assuming each star is an independent tracer, the full likelihood for many stars is then just the product of the individual likelihoods:
%\begin{multline}
%	p(\{\bs{\Theta}^{(i)}\}, \bs{\Theta}_\sat, \bs{\Theta}_\Phi \given \{\D^{(i)}\}, \D_\sat) = \\
%		\frac{1}{\mathcal{Z}} \, \prod_i \, p(\D^{(i)} \given \W^{(i)}) \, p(\D_\sat \given \W_\sat)\,
%			p(\bs{\Theta}^{(i)} \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi) \, 
%			p(\bs{\Theta}) \, p(\bs{\Theta}_\sat)\,p(\bs{\Theta}_\Phi)
%\end{multline}
\begin{multline}
	p(\{\D^{(i)}\}, \D_\sat \given \{\bs{\Theta}^{(i)}\}, \bs{\Theta}_\sat, \bs{\Theta}_\Phi ) = \\
		p(\D_\sat \given \W_\sat) \, \prod_i \, p(\D^{(i)} \given \W^{(i)}) \,
			p(\bs{\Theta}^{(i)} \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi)
\end{multline}
All parameters are summarized in Table~\ref{tbl:params}.

@@

We first test \rewinder\ using only eight stars (particles) --- four from the leading tail, and four from the trailing tail --- randomly sampled from the $2.5\times10^8\,\msun$ simulation (Section~\ref{sec:sims}), assuming the observed 6D positions for both the stars and the progenitor have negligible uncertainties. The stars were required to have been stripped after the first pericentric passage and have a present-day distance within $50$ kpc of the Sun. Figure~\ref{fig:sims} (second column from right) shows the eight randomly chosen stars (black crosses) in Galactic coordinates, over-plotted on all other simulated particles (grey points).

@@

\subsection{Data with near-future uncertainties}\label{sec:exp2}
We next take the same eight stars used in the previous experiment and ``observe'' them with optimistic observational uncertainties. We require these stars to be RR Lyrae variables which are known to be excellent distance indicators via the mid-infrared period-luminosity relation \citep[as shown in, e.g.,][]{madore12}. Nearly 100 RR Lyrae associated with the Sgr stream and $\sim$30 associated with the Orphan stream will be observed with \spitzer as part of the SMASH survey \citep{smashprop} with expected fractional distance uncertainties around $\sim$2\%. These stars will also be included in the \gaia\, proper motion catalog: at a distance of $\sim$50~kpc, a typical RR Lyrae will have a tangential velocity error around $\sim$20~km/s.

For this experiment, we (1) assume the stars are bright distance indicators --- RR Lyrae --- such that the fractional distance uncertainty is $2\%$; (2) neglect the uncertainty in angular position (for a typical RR Lyrae at 50~kpc this is $\sim$$10^{-7}$~deg for \gaia); (3) assume we can measure radial velocities to these stars with 5~km/s uncertainty; and (4) compute the sky-averaged \gaia\, proper-motion uncertainty for each star assuming an F0V spectral type using the \texttt{PyGaia}\footnote{https://github.com/agabrown/PyGaia} code and use this uncertainty for both components of proper motion. We further assume that we know the tail assignment for each star, $\beta$. We observe the position of the progenitor with the same observational uncertainties though in reality some coordinates will have even better measurements.

This experiment samples over the four potential parameters, the Lagrange point offset, four phase-space coordinate parameters plus the unbinding time for each star (40 total), and four phase-space coordinate parameters for the progenitor --- 49 parameters in total. We use an ensemble of 256 walkers and draw initial conditions for the coordinate parameters by sampling from Gaussian's centered on the ``observed'' value with variances specified by the observational uncertainties. Other parameters --- potential parameters, $\Loffset$, and $\tub$ --- are initialized by drawing from the priors summarized in Table~\ref{tbl:params}. [TODO: start from truth...] We burn in the walkers for 10000 steps and run for 10000 inference steps. Figures~\ref{fig:exp2_potential}, \ref{fig:exp2_satellite}, \ref{fig:exp2_particle} show the marginalized posteriors for the potential parameters, satellite parameters, and parameters for a single particle.

% TODO: slight bias in values due to particular sample of stars?

@@

\subsection{Precise distance measurements with missing proper motions}\label{sec:exp3}
The SMASH survey \citep{smashprop} will be completed within a year (early 2015), long before the final \gaia\, data release. Thus, we will soon have precise distance measurements for stars in the Sgr and Orphan streams, but poor proper motion constraints. SMASH also targets several RR Lyrae in the Sgr core, which enable a high-precision distance measurement of the progenitor. Though proper motions have been measured for the Sgr progenitor, we now consider a case in which we have high-precision (2\%) distances to stream stars and the progenitor, 10~km/s radial velocity uncertainty, and missing proper motions for all stars and progenitor.

As with the previous experiment, this experiment includes 49 model parameters in total. We again use an ensemble of 256 walkers, but this time initialize the walkers by sampling the coordinate parameters from Gaussians centered on the true positions with variances set to 1\% of the parameter values. Other parameters --- potential parameters, $\Loffset$, and $\tub$ --- are initialized by drawing from the priors summarized in Table~\ref{tbl:params}. We again burn in the walkers for 10000 steps and run for 10000 inference steps. Figures~\ref{fig:exp3_potential}, \ref{fig:exp3_satellite}, \ref{fig:exp3_particle} show the marginalized posteriors for the potential parameters, satellite parameters, and parameters for a single particle. Accounting for observational uncertainties does increase the uncertainties in the constraints on the potential parameters (as expected), but this demonstrates that a small sample of well-measured stream stars provide an enormous amount of information about the triaxiality of the potential.

@@

\rewinder works well when all stars are lost due to tidal pruning. Even on a mildly eccentric orbit, stars are preferentially stripped around the satellite's pericenter because of tidal shocking. We will explore properly modeling this process in future work. We have focused on modeling streams of stars formed from a disrupting satellite on a mildly eccentric orbit. On more eccentric orbits, the debris tends to form more shell or cloud-like structures. The morphological distinction between stellar streams and stellar clouds is just related to the orbit of the disrupting satellite within the parent galaxy. With a proper model for the tidal shocking that occurs on an eccentric orbit, \rewinder\ could also be applied to structures such as the Triangulum-Andromeda \citep{rochapinto04} or Hercules-Aquila \citep{belokurov07b} stellar clouds, both possible shell-like structures viewed from the inside.

@@

>>>>

\begin{abstract}
% Context
The dark-matter halo of the Milky Way is expected to be triaxial on large scales
  and filled with substructure.
It is hoped that streams or shells of stars produced by tidal disruption
  of smaller galaxies or globular clusters will provide very precise measures of 
  the gravitational potential, 
  to test these predictions.
% Aims
We develop a statistically justified method for inferring the gravitational potential
  using precise measurements of stars in tidal streams
  based on the idea that these stars
  were close in phase space at some point in the past.
Our method is flexible enough to adapt to any form for the Galactic potential.
In particular, it works in phase-space rather than action-space and hence relies neither on
our ability to derive actions nor on the 
assumption of integrability of the potential;
  we do not know that any good approximation to the large-scale 
  Milky Way potential is integrable.
% Methods: 
Our model is probabilistic, with a likelihood function and priors on the parameters.
  These parameters include the present-day phase-space position for each star
  and the progenitor, and thus the method can properly account for both finite observational uncertainties 
  and missing data dimensions.
We test our method on synthetic data sets generated from an N-body simulation
of satellite disruption in a multi-component Milky Way including a triaxial dark matter halo. 
The data sets were generated with
  observational uncertainties to mimic current and near-future surveys of various stars.
% Results:
We find that even with four well-measured stream stars,
  we can infer properties of the triaxial potential ???? with accuracies of order ??? percent ??? and
  missing phase-space coordinates of the stars ??? with accuracies of order ????.
The method generalizes to encompass mixtures of independent debris structures,
  or situations in which star membership in structures is unknown.
% [Get more quantitative here?]
% Conclusions: 
\end{abstract}

@@

@@

Motivated by the strengths of previous work, we identify a minimum set of issues that any approximate stream model or potential recovery method should address (see Section~\ref{sec:discussion} for a more detailed discussion of these points in the context of this work):
\begin{enumerate}
	\item \textbf{Observational uncertainties:} The known debris structures are $\gtrsim$10 kpc from the Sun, where distance and proper motion measurement errors are significant. Thus it is critical for any method that uses tidal debris to incorporate observational uncertainties and missing dimensions in a consistent and justified way. 
	\item \textbf{Form of the potential:} There is large uncertainty in the radial profile, shape, orientation, and graininess of the outer halo and the constancy of these parameters over distance. Properly describing the potential may require non-parametric techniques or complex analytic functions so the method should not rely on the existence of or ability to compute simple conserved orbital properties, such as actions.
	\item \textbf{Multiple debris structures:} Near-future photometric surveys such as \gaia\, and the \project{LSST} will likely discover many new streams and kinematic associations of stars. Potential recovery methods should be able to simultaneously use multiple streams and incorporate other dynamical constraints.
	\item \textbf{Comparing models to data:} Matching generated streams to observed stellar densities is difficult; models that rely on this must account for observational biases, internal properties of the progenitor, and background halo stars. [TODO: are we matching densities?]
	\item \textbf{Computational expense:} Full N-body simulations are expensive to run; incorporating an N-body simulation into a likelihood function evaluation and then performing a parameter search would be computationally intensive and, presently, intractable.
\end{enumerate}

@@

In all simulations, the host potential is taken to be a three-component sum of a Miyamoto-Nagai disk \citep{miyamoto75}, Hernquist spheroid, and a triaxial, logarithmic halo \citep[e.g.,][]{law10}:
\begin{align}
	&\Phi_{\rm disk}(R,z) = -\frac{GM_{\rm disk}}{\sqrt{R^2 + (a + \sqrt{z^2 + b^2})^2}}\\
	&\Phi_{\rm spher}(r) = -\frac{GM_{\rm spher}}{r + c}\\
	&\Phi_{\rm halo}(x,y,z) = \vhalo^2 \ln(C_1 x^2 + C_2 y^2 + C_3 xy + (z/q_z)^2 + r_h^2)
\end{align}
where $C_1$, $C_2$, and $C_3$ are combinations of the $x$ and $y$ axis
ratios ($q_1$, $q_2$) and orientation of the halo with respect to the
baryonic disk ($\phi$):
\begin{align}
  C_1 &= \frac{\cos^2\phi}{q_1^2} + \frac{\sin^2\phi}{q_2^2}\\
  C_2 &= \frac{\sin^2\phi}{q_1^2} + \frac{\cos^2\phi}{q_2^2}\\
  C_3 &= 2\sin\phi\cos\phi \left(q_1^{-2} - q_2^{-2}\right).
\end{align}
The total potential then is just
\begin{equation}
	\Phi_{\rm tot} = \Phi_{\rm disk} + \Phi_{\rm spher} + \Phi_{\rm halo}\label{eq:lm10}.
\end{equation}
Figure~\ref{fig:potential} shows equipotential slices in the Galactic X-Z plane (Y=0) and Y-Z plane (X=0) for a few choices of  $q_1$,  $q_z$, and $\phi$ while holding all other parameters fixed, as described in the figure caption. The potential parameters used for the simulations are shown in the ``Truth'' column of Table~\ref{tbl:params}.

@@

In a triaxial potential, the orbital plane of the satellite is not fixed but we still expect there to be \emph{effective} Lagrange points for a spherical satellite system along the line of centers connecting the origin of the parent potential to the satellite --- that is, we expect the stars to be stripped at some characteristic radius near the tidal radius with some dispersion about this point. We proceed by defining a coordinate system relative to the position and velocity of the progenitor, ($\bs{r}_p$,$\bs{v}_p$), rotated into the instantaneous orbital plane defined by $\bs{L} = \bs{r}_p \times \bs{v}_p$. The (time-dependent) basis vectors are given by
\begin{align}
	\hat{\bs{x}}_1 &= \frac{\bs{r}_p}{\|\bs{r}_p\|}\label{eq:x1}\\
	\hat{\bs{x}}_2 &= \frac{\bs{L} \times \hat{\bs{x}}_1}{\|\bs{L} \times \hat{\bs{x}}_1\|} = \hat{\bs{x}}_3 \times \hat{\bs{x}}_1\\
	\hat{\bs{x}}_3 &= \hat{\bs{L}} = \frac{\bs{L}}{\|\bs{L}\|} = \frac{\bs{r}_p \times \bs{v}_p}{\|\bs{r}_p \times \bs{v}_p\|}\label{eq:x3}.
\end{align}
Figure~\ref{fig:lpts_r} shows sections of particle orbits (for particles that are stripped) from the N-body simulations described above, projected into this coordinate system. The positions are normalized by the instantaneous tidal radius (Eq.~\ref{eq:tidalradius}) and velocities are normalized by the instantaneous velocity scale (Eq.~\ref{eq:velscale}). Sections of the orbits symmetric around their unbinding times, $\tub$, are shown for each of the four progenitor masses. In these coordinates, the classical Lagrange points would be located at $x_1\approx\pm\rtide,x_2=0,x_3=0$ (illustrated by the intersection of the black unit circles and horizontal lines in Figure~\ref{fig:lpts_r}). Though not centered on the point-mass Lagrange points, the location of and dispersion about the effective Lagrange points in these scaled coordinates remain remarkably consistent across the range of progenitor masses explored. Figure~\ref{fig:lpts_v} shows the velocity orbits of each star also projected into these coordinates and normalized. The dispersion in velocity is well-normalized by the velocity scale of Eq.~\ref{eq:velscale}, though it is clear that the velocity dispersion along $\hat{x}_1$, the radial vector to the satellite position, is larger than that in other dimensions, possibly because of tidal shocking.

@@

To model the true, 6D, present-day position of the star, $\W$, we first transform to cartesian, Galactocentric coordinates where ($\bs{r}_0,\bs{v}_0$) and ($\bs{r}_{\sat,0},\bs{v}_{\sat,0}$) are the position and velocity of the star and progenitor today. The star is taken as having been sampled at $t=\tub$ (the unbinding time) from an isotropic Gaussian centered on one of the effective Lagrange points in position, and an isotropic Gaussian centered on the origin in velocity. The present-day phase-space position is the result of integrating this sample forward from $\tub$ in the parent potential, $\Phi$, whose form is parametrized by the vector $\bs{\Theta}_\Phi$. We assume that once the star becomes ``unbound'' from the satellite the potential of the satellite can be ignored,\footnote{This assumption breaks down for sufficiently large-mass progenitors, somewhere between $\sim10^8-10^9~\msun$.} and thus treat the star as a test particle. The satellite mass enters this prescription through the tidal radius; a changing satellite mass will skew the positions of the effective Lagrange points and velocity scale. We allow for mass-loss by incorporating the initial satellite mass, $m_0$, and a constant mass-loss term, $\dot{m}$ into the model: $m(t) = m_0 - \dot{m}t$. Finally, we add two additional parameters: a constant, global factor, $\Loffset$, that scales the position of the effective Lagrange points relative to the classical tidal radius, and a binary parameter $\tailbit$, that is either $-1$ or $+1$ depending on whether the star is in the leading or trailing tail. To compress notation, we pack all progenitor parameters ($\W_\sat, m_0, \dot{m}, \Loffset$) into the vector $\bs{\Theta}_\sat$. The likelihood is then: %, and all star parameters ($\W, \tailbit, \tub$) into the vector $\bs{\Theta}$
\begin{align}
	p(\W \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi, \tailbit, \tub) &= p(\bs{r}_0,\bs{v}_0, \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi, \tailbit, \tub)\,\left\vert\J\right\vert\label{eq:starlike}\\
	p(\bs{r}_0,\bs{v}_0, \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi) &= \mathcal{N}(\bs{r} \given \bs{r}_\sat - \Loffset\,\tailbit\,\rtide\,\hat{x}_1, \rtide^2)\,\mathcal{N}(\bs{v} \given \bs{v}_\sat, \sigma^2_v)
\end{align}
where $\rtide$ is given by Eq.~\ref{eq:tidalradius}, and $\sigma_v$ by Eq.~\ref{eq:velscale} and the positions and velocities ($\bs{r},\bs{v}, \bs{r}_{\sat},\bs{v}_{\sat}$) are evaluated at the unbinding time, $\tub$, by integrating the orbits backwards from ($\bs{r}_0,\bs{v}_0, \bs{r}_{\sat,0},\bs{v}_{\sat,0}$). $\left\vert\J\right\vert = \left\vert\frac{\partial(x,y,z,v_x,v_y,v_z)}{\partial(l,b,d,\mu_l,\mu_b,v_r)}\right\vert$ is the absolute value of the determinant of the Jacobian that defines the transformation from heliocentric, spherical to Galactocentric, cartesian coordinates.

The likelihood above is evaluated for each star at its unbinding time, however this is not a priori known for any star. Furthermore, the idea of a single-valued unbinding time is not well defined as (1) it depends strongly on the particular definition of ``unbound'' and (2) for debris disrupted on near-circular orbits, the debris may pass close to the progenitor several times. We marginalize over the unbinding time, assuming a uniform distribution of unbinding times over the entire interaction time:
\begin{align}
	p(\W \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi, \tailbit) &= \int p(\W \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi, \tailbit, \tub)\,p(\tub)\,d\tub
\end{align}

@@

For a single star, the evaluation of the likelihood works as follows: given present-day coordinates for the star and progenitor, potential parameter values ($\bs{\Theta}_\Phi$), and nuisance parameter values ($\tub,\Loffset,\tailbit$),
\begin{enumerate}
	\item transform star and progenitor positions in heliocentric coordinates to Galactocentric, cartesian coordinates;
	\item integrate star and progenitor orbits backwards as test particles in the potential given by $\bs{\Theta}_\Phi$ for the total interaction time, $t_{\rm int}\approx6$~Gyr;
	\item transform the particle position to the relative, normalized coordinates defined in Section~\ref{sec:debris};
	\item compute the likelihood given by Equation~\ref{eq:starlike} at each timestep of the integration and marginalize over the unbinding time, $\tub$.
\end{enumerate}

Assuming each star is an independent tracer, the full likelihood for many stars is then just the product of the individual likelihoods:
%\begin{multline}
%	p(\{\bs{\Theta}^{(i)}\}, \bs{\Theta}_\sat, \bs{\Theta}_\Phi \given \{\D^{(i)}\}, \D_\sat) = \\
%		\frac{1}{\mathcal{Z}} \, \prod_i \, p(\D^{(i)} \given \W^{(i)}) \, p(\D_\sat \given \W_\sat)\,
%			p(\bs{\Theta}^{(i)} \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi) \, 
%			p(\bs{\Theta}) \, p(\bs{\Theta}_\sat)\,p(\bs{\Theta}_\Phi)
%\end{multline}
\begin{multline}
	p(\{\D^{(i)}\}, \D_\sat \given \{\bs{\Theta}^{(i)}\}, \bs{\Theta}_\sat, \bs{\Theta}_\Phi ) = \\
		p(\D_\sat \given \W_\sat) \, \prod_i \, p(\D^{(i)} \given \W^{(i)}) \,
			p(\W^{(i)} \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi, \tailbit^{(i)})
\end{multline}
where $\bs{\Theta}^{(i)} = (\W^{(i)}, \tailbit^{(i)})$. All parameters are summarized in Table~\ref{tbl:params}.

@@

We first test \rewinder\ using only four stars (particles) --- two from the leading tail, and two from the trailing tail --- randomly sampled from the $2.5\times10^8\,\msun$ simulation (Section~\ref{sec:sims}), assuming the observed 6D positions for both the stars and the progenitor have negligible uncertainties. The stars were required to have been stripped after the first pericentric passage and have a present-day distance within $50$ kpc of the Sun. Figure~\ref{fig:sims} (second column from right) shows the eight randomly chosen stars (black crosses) in Galactic coordinates, over-plotted on all other simulated particles (grey points).

@@

\subsection{Data with near-future uncertainties}\label{sec:exp2}
We next take the same four stars used in the previous experiment and ``observe'' them with optimistic observational uncertainties. We require these stars to be RR Lyrae variables which are known to be excellent distance indicators via the mid-infrared period-luminosity relation \citep[as shown in, e.g.,][]{madore12}. Nearly 100 RR Lyrae associated with the Sgr stream and $\sim$30 associated with the Orphan stream will be observed with \spitzer as part of the SMASH survey \citep{smashprop} with expected fractional distance uncertainties around $\sim$2\%. These stars will also be included in the \gaia\, proper motion catalog: at a distance of $\sim$50~kpc, a typical RR Lyrae will have a tangential velocity error around $\sim$20~km/s.

For this experiment, we (1) assume the stars are RR Lyrae stars (e.g., bright distance indicators) such that the fractional distance uncertainty is $2\%$; (2) neglect the uncertainty in angular position, $l,b$ (for a typical RR Lyrae at 50~kpc this is $\sim$$10^{-7}$~deg for \gaia); (3) assume we can measure radial velocities to these stars with 5~km/s uncertainty; and (4) compute the sky-averaged \gaia\, proper-motion uncertainty for each star assuming an F0V spectral type using the \texttt{PyGaia}\footnote{https://github.com/agabrown/PyGaia} code and use this uncertainty for both components of proper motion. We further assume that we know the tail assignment for each star, $\beta$. We observe the position of the progenitor with the same observational uncertainties though in reality some coordinates will have even better measurements.

This experiment samples over the four potential parameters, the Lagrange point offset, four phase-space coordinate parameters for each star (16 total), and four phase-space coordinate parameters for the progenitor --- 25 parameters in total. We use an ensemble of 256 walkers and draw initial conditions for the coordinate parameters by sampling from Gaussian's centered on the ``observed'' value with variances specified by the observational uncertainties. Other parameters --- potential parameters and $\Loffset$ --- are initialized by drawing from the priors summarized in Table~\ref{tbl:params}. We burn in the walkers for 50000 steps and run for 100000 inference steps, but thin the chains by taking every $2\times \max(t_{\rm acor})$ sample, where $\max(t_{\rm acor}) \approx 4000$~steps. Figures~\ref{fig:exp2_potential}, \ref{fig:exp2_satellite}, \ref{fig:exp2_particle} show the marginalized posteriors for the potential parameters, satellite parameters, and parameters for a single particle.

@@

\subsection{Precise distance measurements with missing proper motions}\label{sec:exp3}
The SMASH survey \citep{smashprop} will be completed within a year (early 2015), long before the final \gaia\, data release. Thus, we will soon have precise distance measurements for stars in the Sgr and Orphan streams, but poor or no proper motion constraints. SMASH also targets several RR Lyrae in the Sgr core, which will enable a high-precision distance measurement of the progenitor. Though proper motions have been measured for the Sgr progenitor, we now consider a case in which we have high-precision (2\%) distances to stream stars and the progenitor, 10~km/s radial velocity uncertainty, but missing proper motions for all stars and progenitor.

As with the previous experiment, this experiment includes 25 model parameters in total. We again use an ensemble of 256 walkers. Other parameters --- potential parameters and $\Loffset$ --- are initialized by drawing from the priors summarized in Table~\ref{tbl:params}. We burn in the walkers for 50000 steps and run for 500000 inference steps. We again thin the chains by taking every $2\times \max(t_{\rm acor})$ sample, but here the autocorrelation time is found to be very long, $\max(t_{\rm acor}) \approx 25000$~steps. Figures~\ref{fig:exp3_potential}, \ref{fig:exp3_satellite}, \ref{fig:exp3_particle} show the marginalized posteriors for the potential parameters, satellite parameters, and parameters for a single particle. Even a small sample of stars with precise distance measurements provide an enormous amount of information about the triaxiality of the potential.

@@

\rewinder\, works well when all stars are lost due to tidal pruning. Even on a mildly eccentric orbit, stars are preferentially stripped around the satellite's pericenter because of tidal shocking. We will explore properly modeling this process in future work. We have focused on modeling streams of stars formed from a disrupting satellite on a mildly eccentric orbit. On more eccentric orbits, the debris tends to form more shell or cloud-like structures. The morphological distinction between stellar streams and stellar clouds is just related to the orbit of the disrupting satellite within the parent galaxy. With a proper model for the tidal shocking that occurs on an eccentric orbit, \rewinder\ could also be applied to structures such as the Triangulum-Andromeda \citep{rochapinto04} or Hercules-Aquila \citep{belokurov07b} stellar clouds, both possible shell-like structures viewed from the inside.

@@

% TODO: We have tested leaving the unbinding time as a free parameter but have found that this makes the likelihood severely multi-modal and thus very difficult to sample
